---
 title: "LTMLE manual"
 output: pdf_document
 date: "08-01-2023"
---

The aim of this document is to provide practical advice on how to structure typical observational data in order to conduct a longitudinal targeted maximum likelihood estimation (LTMLE) analysis.

Note that multiple outcomes can be managed simultaneously and multiple time periods can be managed by selecting the maximum time period.
 
This practical manual is based on the data.table package and further the heaven package which is available on github/tagteam/heaven.

The manual is also hopefully based on a typical presentation of raw data.  

# Raw data formatting
First we will create the basic data structure necessary for the analysis.
All basic information of time dependent variables is based on dates of events. The following block creates the data for this example

```{r echo=TRUE}

#install.packages(c('data.table','devtools')) # install 'data.table' and 'devtools' if needed
library(devtools)
library(data.table)
#install_github('tagteam/heaven') # install 'heaven' if needed
#library(heaven)
library(ltmle)

set.seed(21)
n <- 30
W <- rnorm(n) # A baseline variable
ID<-1:n
startDate<-as.Date("2010-01-01")+sample(0:365,n,replace=TRUE)
L1<-startDate+sample(0:365,n,replace=TRUE)
L2<-startDate+sample(0:365,n,replace=TRUE)
ebd <- pmax(L1,L2)+sample(0:180,n,replace=TRUE)
base <- data.table(ID,startDate)
fixed <- data.table(ID,W)
A <- sample(0:1,n,replace=TRUE)
select <- sample(1:3,n,replace=TRUE) # event, censor, competing
covariatesEvents <- data.table(ID,A,L1,L2,select,ebd)
covariatesEvents[select==3 | A==1,Y:=ebd] # A is really dangerous
covariatesEvents[select==2,C:=ebd]
covariatesEvents[select==1,compete:=ebd]
covariatesEvents[,c("select","ebd","A"):=NULL]
select2 <- sample(1:2,n,replace=TRUE)
select3 <- sample(1:2,n,replace=TRUE)

L3 <- sample(0:1,n,replace=TRUE)
timeVarCov <- data.table(ID,A,L3,select2,startDate)
timeVarCov2 <- timeVarCov[A==1]
timeVarCov2[,':='(inTime=startDate,outTime=startDate+180+sample(0:365,1),var='A')]
timeVarCov3 <- timeVarCov[L3==1]
timeVarCov3[,':='(inTime=startDate,outTime=startDate+sample(366:720,1),var='L3')]
timeVarCov <- rbind(timeVarCov2,timeVarCov3)
timeVarCov <- timeVarCov[!is.na(inTime),.(ID,inTime,outTime,var)]
timeVarCov[,value:='1']
```

To use this manual, the raw data should be formatted as four datasets. The
first contains the individual identification (ID) and the date of entry
into the study (startDate). 

```{r echo=TRUE}
base
```
The second dataset includes the ID and fixed baseline covariate (W). Note that 
baseline covariates can be both continuous and discrete whereas all time dependent
variables need to be discrete. For the current example there is a single continuous
baseline variable (W), but there can be any number.

```{r echo=TRUE}
fixed
```

The third dataset includes dates of variables that only change once
during follow-up. This includes covariates such as diseases (L) and information 
about when and how the follow-up ends in either outcome (Y), censoring (C), and competing
event (compete). The dataset to be presented includes these variables and the value
of all variables is either the date of the change or missing (NA).

```{r echo=TRUE}
covariatesEvents
```

The fourth dataset includes data for variables that may change multiple times.
For the present example it includes a third covariate (L3) and the treatment of interest (A). 
This dataset includes the identification (ID), a variable (var) where the levels 
correspond to the names of the time varying variable (i.e. 'L3' and 'A') as well
as start and end of each period (inTime/outTime) and a variable (value) with levels
"1" indicating "exposure" and "0" indicating "no exposure".

```{r echo=TRUE}
timeVarCov
```

# Splitting by time varying events

LTMLE operates with a sequence of time periods all of equal length. Even though 
individuals may experience the event or be censored early, all individuals need
information for variables during the whole time of follow-up. Therefore, the first
step is to fix the end time of the analysis. 

For this example, we chose four time intervals of 180 days.

During the data preparation below, records are split into multiple records and 
to avoid confusion with the original entry into analysis, a new variable is 
defined informing the start date.

```{r echo=TRUE}
base[,':='(inn=startDate,out=startDate+4*180)]
```

The following steps have the purpose of defining levels of variables in each of 
the defined time periods. To start this proces all records are split according 
to timing of change in variable status. The order of splitting is not important.

The first step is splitting by all variables that change only once. 
This is performed with the heaven::splitTwo function that needs the original 
base data and a "splitting guide" which is the dataset with dates where variables
that change only once are held.

```{r echo=TRUE}
longSplit <- splitTwo(indat=base,
                      splitdat=covariatesEvents,
                      invars=c('ID','inn','out'),
                      splitvars = c('ID',"L1","L2","Y","C","compete"))
```

Next the data is split by the time dependent variables with potentially multiple
changes during the study. We use the function heaven::splitFromTo.

```{r echo=TRUE}
longSplit <- splitFromTo(indat=longSplit,
                         splitdat=timeVarCov,
                         invars=c('ID','inn','out'),
                         splitvars = c('ID','inTime','outTime','value','var'))
```

Finally, the data is split by the selected time periods, in the current case
four periods of 180 days. The new value 'period' contains the period number. This uses
heaven::lexisSeq.

```{r echo=TRUE}
longSplit <- splitSeq(indat=longSplit,
                      invars=c('ID','inn','out'),
                      varname = 'startDate', # intervals since inclusion in study
                      splitvector= c(-1,180,360,540,720), # four periods of 180 days
                      format = "vector",
                      value="period")
```

# Aggregate in time periods

With the splitting complete, all information for each selected equally sized time period 
(in the example four) is separated.

The next step is then to summarize information by 'period'. 

The outcomes (event, censoring, competing risk) should be the maximal outcome for each period since an outcome event is coded "1" as opposed to "0" when not occurring.  

For other variables, it is dependent on a discussion whether to choose the entry values
for each period or choose any indication of exposure during the period. For this
example the entry value is chosen for each period for covariates. 

Importantly, the data structure for running LTMLE requires that when an event has
happened during a period, the value of that event needs to be carried forward to the
rest of periods until end of observation time.  

This last step is performed based on experience with the package, and appears 
to be in contrast with the LTMLE manual where any value can appear later.

For the calculations, the covariates (as opposed to outcomes) need to be moved 
one period back in time. This ensures that covariate values are always prior to outcome. 
Conceptually covariate values can be viewed as the last value during the preceding time period.


```{r echo=TRUE}
setkeyv(longSplit,c("ID","period","inn"))
# Max value of outcomes in each period
longSplit[,':='(Y=max(Y),C=max(C),compete=max(compete)),by=c("ID","period")] 
# Carry outcomes forward until end of observation
setkeyv(longSplit,c("ID","inn"))
longSplit[shift(Y)==1 & shift(ID)==ID,Y:=1]
longSplit[shift(C)==1 & shift(ID)==ID,C:=1]
longSplit[shift(compete)==1 & shift(ID)==ID,compete:=1]
# Choose first record for each ID/period
aggrSplit <- longSplit[,.SD[1],by=c("ID","period")]
aggrSplit[,A:=as.numeric(A)]
aggrSplit[,L3:=as.numeric(L3)]
# shift covariates one period back 
aggrSplit_cov <- aggrSplit[,.(ID,period,A,L3,L1,L2)]
aggrSplit_cov[,period:=period-1]
aggrSplit_out <- aggrSplit[,.(ID,period,compete,C,Y)]
```

# Summarizing presence of data

The positivity assumption requires that the propensity of the covariate presence 
is above zero and below one in each period. 
A preliminary test is simply to tabulate the presence of covariates in each
period.

```{r echo=TRUE}
aggrSplit_cov[,.(L1=sum(L1)/length(L1),L2=sum(L2)/length(L2)),by="period"]
```

# Transpose to wide format
```{r echo=TRUE}
wideSplit_cov <- dcast(aggrSplit_cov,ID~period,value.var = c("A", "L1","L2","L3"))
wideSplit_out <- dcast(aggrSplit_out,ID~period,value.var = c("compete","C","Y"))
wideSplit <- cbind(wideSplit_cov,wideSplit_out[,ID:=NULL])
# Finally add the baseline covariates
wideSplit <- merge(fixed,wideSplit,by="ID")
```

# Deterministic Q-function

The deterministic Q function can be employed for a variety of purposes. Centrally 
for our example, it interprets the "compete" covariate as a competing risk.

```{r, echo=T}
det.Q.function<-function(data, current.node, nodes, called.from.estimate.g){
    compete.index <- grep("compete",names(data))
    ## if(length(compete.index)==0) det <- list(rep(TRUE,NROW(data)),Q.value=0)
    ## stop("No compete/terminal event node found")
    hist.compete.index <- compete.index[compete.index < current.node]
    if(length(hist.compete.index)==0)
      return(NULL)
    else{
      is.deterministic <- Reduce("+",lapply(data[,hist.compete.index,drop=FALSE],
                                            function(dd){x=dd;x[is.na(dd)] <- 0;x}))>=1
      is.deterministic[is.na(is.deterministic)] <- FALSE
      list(is.deterministic=is.deterministic, Q.value=0)
    }
}
    
```

#LTMLE
What is left is to run the LTMLE analysis, and more information on this function exists: 
https://cran.r-project.org/web/packages/ltmle/ltmle.pdf

```{r echo=TRUE, message=F,warning=F}
Anodes <- sort(grep("^A",names(wideSplit)))
Lnodes <- sort(c(grep("^L",names(wideSplit)),grep("^W",names(wideSplit)),grep("^compete_",names(wideSplit))))
Ynodes <- sort(grep("^Y",names(wideSplit)))
Cnodes <- sort(grep("^C",names(wideSplit)))
    
fit <- ltmle(wideSplit,Anodes=Anodes, 
          Lnodes=Lnodes, 
          Ynodes=Ynodes,
          Cnodes=Cnodes,
          abar=c(1,1,1,1),
          deterministic.Q.function = det.Q.function, #defined above
          survivalOutcome = TRUE)

print(summary(fit))
    ```

[Example of interpretation of estimates?]

<!-- # LTMLE -->
<!-- ```{r echo=TRUE} -->
<!-- # Order variables -->
<!-- W <- grep("^W",names(wideSplit)) -->
<!-- L <- grep("^L",names(wideSplit)) -->
<!-- A <- grep("^A",names(wideSplit)) -->
<!-- Y <- grep("^Y",names(wideSplit)) -->
<!-- C <- grep("^C",names(wideSplit)) -->
<!-- compete <- grep("^compete",names(wideSplit)) -->
<!-- neworder <- c() -->
<!-- setcolorder(wideSplit,names(wideSplit)[c(1,compete,L,W,A,Y,C)]) -->

<!-- Anodes <- grep("^A",names(wideSplit)) -->
<!-- Lnodes <- grep("^L",names(wideSplit)) -->
<!-- Ynodes <- grep("^Y",names(wideSplit)) -->
<!-- Cnodes <- grep("^C",names(wideSplit)) -->
<!-- ltmle(wideSplit,Anodes=Anodes, Lnodes=Lnodes, Ynodes=Ynodes,Cnodes=Cnodes,abar=c(1,1,1,1), -->
<!--       survivalOutcome = TRUE) -->
<!-- ``` -->
