---
 title: "LTMLE manual"
 output: pdf_document
 date: "08-01-2023"
---

The aim of this document is to provide practical advice on how to structure typical observational data in order to conduct a longitudinal targeted maximum likelihood estimation (LTMLE) analysis.

Not that multiple outcomes can be managed simultaneously and multiple time periods can be managed by selecting the maximum time period.
 
This practical manual is based on the data.table package and further the heaven package which is available on github/tagteam/heaven.

# Raw data formatting
First we will create the basic data necessary for the analysis as they often are available.
All basic information of time dependent variables is based on dates of events. The following block creates the data for this example

```{r echo=TRUE}

#install.packages(c('data.table','devtools')) # install 'data.table' and 'devtools' if needed
library(devtools)
library(data.table)
#install_github('tagteam/heaven') # install 'heaven' if needed
library(heaven)

set.seed(21)
n <- 1000
W <- rnorm(n) # A baseline variable
ID<-1:n
startDate<-as.Date("2010-01-01")+sample(0:365,n,replace=TRUE)
L1<-startDate+sample(0:365,n,replace=TRUE)
L2<-startDate+sample(0:365,n,replace=TRUE)
ebd <- pmax(L1,L2)+sample(0:180,n,replace=TRUE)
base <- data.table(ID,startDate,dummy=1,W)
select <- sample(1:3,n,replace=TRUE) # event, censor, competing
covariatesEvents <- data.table(ID,L1,L2,select,ebd)
covariatesEvents[select==3,Y:=ebd]
covariatesEvents[select==2,C:=ebd]
covariatesEvents[select==1,compete:=ebd]
covariatesEvents[,c("select","ebd"):=NULL]
select2 <- sample(1:2,n,replace=TRUE)
select3 <- sample(1:2,n,replace=TRUE)
timeVarCov <- data.table(ID,select2,startDate)
timeVarCov[select2==1,':='(inTime=startDate,outTime=startDate+sample(0:720,1),var='L3')]
timeVarCov[select2==2,':='(inTime=startDate+75,outTime=startDate+75+sample(0:720,1))]
timeVarCov2 <- data.table(ID,select3,startDate)
timeVarCov2[select3==1,':='(inTime=startDate,outTime=startDate+sample(180:720,1),var='A')]
timeVarCov<-rbind(timeVarCov,timeVarCov2,fill=TRUE)
timeVarCov<-timeVarCov[!is.na(inTime) & !is.na(var),.(ID,var,inTime,outTime)]
timeVarCov[,value:='1']
```

To use this manual, the raw data should be formatted as three datasets. The
first contains the individual identification (ID), the date of entry
into the study (startDate), variables that are fixed at baseline (W) 
and a dummy variable that should be either zero or one

```{r echo=TRUE}
base
```

The second dataset includes dates of variables that only change once
during follow-up.  This includes covariates such as diseases (L) and information about  
also how the follow-up ends in either outcome (Y), censoring (C) and competing
event (compete)

```{r echo=TRUE}
covariatesEvents
```

The third dataset includes data for variables that may change multiple times
including multiple times for each individual.  For the present example it includes a
third covariate (L3) and the treatment of interest (A). This dataset includes
the identification (ID), a variable (var) where the levels correspond to the names of the time varying variable (i.e. 'L3' and 'A')
, start and end of each period (inn/out) and a variable (value) with the
content "1" indicating "exposure".

```{r echo=TRUE}
timeVarCov
```

# Splitting by time varying events

The first step is to fix the end time of the analysis. In LTMLE we use a range of 
equally lengthened time intervals and the same number of time intervals is used for all
individuals.

For this example, we chose 4 time intervals of 180 days.

As we split the records it becomes confusing that the original startDate changes, wherefore
a copy of the original startDate is made.

```{r echo=TRUE}
base[,':='(inn=startDate,out=startDate+4*180)]
```


The data are next split by all time dependent variables and events. The order of
splitting is not important.

To split by the covariates and events where each is characterized by a single date 
the heaven::lexisTwo function is used.

```{r echo=TRUE}
longSplit <- lexisTwo(indat=base,
                      splitdat=covariatesEvents,
                      invars=c('ID','inn','out','dummy'),
                      splitvars = c('ID',"L1","L2","Y","C","compete"))
```

Next the data are split by the time dependent variables with potentially multiple
changes during the study. We use the function heaven::lexisFromTo.

```{r echo=TRUE}
longSplit <- lexisFromTo(indat=longSplit,
                         splitdat=timeVarCov,
                         invars=c('ID','inn','out','dummy'),
                         splitvars = c('ID','inTime','outTime','value','var'))
```

Finally, the data are split by the selected time periods, in the current case
four periods of 180 days. The new value 'period' contains the period number. This uses
heaven::lexisSeq.

```{r echo=TRUE}
longSplit <- lexisSeq(indat=longSplit,
                      invars=c('ID','inn','out','dummy'),
                      varname = 'startDate', # intervals since inclusion in study
                      splitvector= c(-1,180,360,540,720), # four periods of 180 days
                      format = "vector",
                      value="period")
```

# Aggregate in time periods

The next step is to summarize information by 'period'. The outcomes (event,
censoring, competing risk) needs to be the mal observation in each period (max is outcome).  
For other variables it is dependent on a discussion whether to choose the entry values
for each period or choose any indication of exposure during the period. For this
example the entry is chosen for each period for covariates and the maximal
for outcomes.

Also when an event has happened during a period the value of that event needs to
be carried forward to the end of observation time.  This last is made from advice 
and appears to be in contrast with the manual where any value can appear later.

```{r echo=TRUE}
setkeyv(longSplit,c("ID","period","inn"))
# Max value of outcomes in each period
longSplit[,':='(Y=max(Y),C=max(C),compete=max(compete)),by=c("ID","period")] 
# Carry outcomes forward until end of observation
setkeyv(longSplit,c("ID","inn"))
longSplit[shift(Y)==1 & shift(ID)==ID,Y:=1]
longSplit[shift(C)==1 & shift(ID)==ID,C:=1]
longSplit[shift(compete)==1 & shift(ID)==ID,compete:=1]
# Choose first record for each ID/period
aggrSplit <- longSplit[,.SD[1],by=c("ID","period")]
aggrSplit[,A:=as.numeric(A)]
aggrSplit[,L3:=as.numeric(L3)]
```
# Positivity table
It is critical for this method that the positivity assumption of causal inference is held.
Thus, for all covariates and all time intervals the probability of covariate presence needs to be
below one and above zero.
```{r echo=TRUE}
aggrSplit[,.(L1=sum(L1)/length(L1),L2=sum(L2)/length(L2)),by="period"]
```

# Transpose to wide format
```{r echo=TRUE}
wideSplit <- dcast(aggrSplit,ID~period,value.var = c("A", "L1","L2","L3","Y", "C", "compete","W") )
```

# LTMLE
```{r echo=TRUE}
# Order variables
W <- grep("^W",names(wideSplit))
L <- grep("^L",names(wideSplit))
A <- grep("^A",names(wideSplit))
Y <- grep("^Y",names(wideSplit))
C <- grep("^C",names(wideSplit))
compete <- grep("^compete",names(wideSplit))
neworder <- c()
setcolorder(wideSplit,names(wideSplit)[c(1,compete,L,W,A,Y,C)])

Anodes <- grep("^A",names(wideSplit))
Lnodes <- grep("^L",names(wideSplit))
Ynodes <- grep("^Y",names(wideSplit))
Cnodes <- grep("^C",names(wideSplit))
ltmle(wideSplit,Anodes=Anodes, Lnodes=Lnodes, Ynodes=Ynodes,Cnodes=Cnodes,abar=c(1,1,1,1),
      survivalOutcome = TRUE)
```
